Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-05-12 doc/invt.tex
small edits to invt.tex

%To simplify the ensuing discussion,
Second, we limit ourselves to matrices % $\mathbf w$'s
that satisfy
the detailed balance condition,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  ,
  \label{eq:w_detailedbalance}
\end{equation}
%
for the error weighting distribution, $\pmb \rho$,
introduced in Eq. \eqref{eq:vbar_def}.
%
This requires the updating matrix, $\mathbf w$
to be symmetric
for a uniform $\pmb\rho$,
or more generally,
the scaled matrix, $\sqrt{ \rho_i/\rho_j } \, w_{ij}$,
to be symmetric.
%
This symmetry allows the diagonalization
of $\mathbf w$ with a set of
eigenvectors, $\phi_{ki}$,
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 1}^n
    \phi_{ki} \, \phi_{kj}
  &=
  \delta_{ij} \, \rho_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li} }
         { \rho_i }
  &=
  \delta_{kl}
  .
  \label{eq:eig_orthonormal_rows}
\end{align}

The above symmetry implies the existence of
a left eigenvector, $\pmb \rho$,
corresponding to the uniform right eigenvector
associated with Eq. \eqref{eq:w_sumj},
with eigenvalue $1$:
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j
  .
  \label{eq:w_balance}
\end{equation}
%
We will label this eigenvector as the first one,
and
%
\begin{equation}
  \phi_{1i} = \rho_i,
\label{eq:eigenmode1}
\end{equation}
%
with $\lambda_1 = 1$.
%
Clearly, Eq. \eqref{eq:eigenmode1}
satisfies the normalization condition
given by Eq. \eqref{eq:eig_orthonormal_rows}
for $k = l = 1$,
while the orthogonality for $k \ge 2$ demands
%
\begin{equation}
  \sum_{ i = 1 }^n \phi_{ki}
  =
  \sum_{ i = 1 }^n
    \frac{ \phi_{ki} \, \phi_{1i} }
         { \rho_i }
  =
  0
  .
\label{eq:ortho1}
\end{equation}
%
We may write the updating matrix in terms of $\phi_{ki}$'s as
%
\begin{equation}
  w_{ij}
  =
  \frac{1}{\rho_i} \sum_{k=1}^n \lambda_k \, \phi_{ki} \, \phi_{kj}
  .
  \notag
\end{equation}

@@

The error defined in Eq. \eqref{eq:error_def}
can be rewritten in terms of the eigenmodes of $\mathbf w$,
%
\begin{align}
  \Err
  \equiv
  \sum_{i = 1}^n \rho_i \,
    \left\langle x_i^2 \right\rangle
  =
  \sum_{i, j, k=1}^n \phi_{ki} \, \phi_{kj} \,
    \left\langle x_i \, x_j \right\rangle
  =
  \sum_{k = 1}^n
    \left\langle
      y_k^2
    \right\rangle
  ,
\notag
\end{align}
%
where we have used Eq. \eqref{eq:eig_orthonormal_cols}.
%
From Eq. \eqref{eq:eigenmode1},
the first eigenmode,
$y_1 = \sum_{j=1}^n \rho_j \, x_j$,
vanishes identically by Eq. \eqref{eq:rhox_sum}.
%
Thus, we can start the last sum %in Eq. \eqref{eq:y2_sum}
from $k = 2$,
%
\begin{equation}
  \Err(T)
  =
  \sum_{k = 2}^n
    \left\langle
    y_k^2(T)
    \right\rangle
  .
\label{eq:y2_sum}
\end{equation}

In analogy to Eq. \eqref{eq:ET_average},
%using Eqs. \eqref{eq:yt_diffeq} and \eqref{eq:y2_sum},
we can write the error after a period, $T$,
as a sum of two contributions,
%
\begin{equation}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T),
  \label{eq:error_tot}
\end{equation}
%
where the residual error
(from the decay of the initial error)
is given by
%
\begin{equation}
  \Err_R(T)
  =
  \sum_{k = 2}^n
    \left\langle y_k^2(0) \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
and the asymptotic error
is given by
%
\begin{align}
  \Err_A(T)
  &=
  \int_0^T
  \sum_{k = 2}^n
  \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr) \, dt
  ,
\label{eq:error_asym}
\end{align}
%
where
%
\begin{equation}
  u_k(q') \equiv e^{\lambda_k \, [q' - q(T)]}
  ,
  \label{eq:uk_def}
\end{equation}
%
and $\Gamma_k$ is computed from
the autocorrelation function
of the histogram fluctuation,
$\kappa_k(t) \equiv \langle \theta_k(0) \, \theta_k(t) \rangle$,
as:
%
\begin{equation}
  \Gamma_k
  = \sum_{t = -\infty}^\infty \kappa_k(t)
  = \sum_{t = -\infty}^\infty \left\langle \theta_k(0) \, \theta_k(t) \right\rangle
  .
  \label{eq:Gamma_sum}
\end{equation}
%
%\begin{equation}
%$\Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt$,
%  \notag
%\label{eq:Gamma_integral}
%\end{equation}
%
Equation \eqref{eq:Gamma_sum}
essentially maps the histogram fluctuation, $\theta_k(t)$,
to the equivalent white noise
so that the theory can be applicable to
a general sampling process.
%
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
For example, we expect large values of $\Gamma_k$
from eigenmodes corresponding to
persistent long-range motion
in sampling complex or glassy systems.
%
In Appendix \ref{sec:Gamma},
we evaluate $\Gamma_k$'s in two special cases,
and show how to measure them
from an adaptive FDS simulation.
%
%\begin{equation}
%  \Err_A(T)
%  =
%  \int_0^T \int_0^T
%  \sum_{k = 2}^n
%    \dot u_k\bigl( q(t) \bigr) \,
%    \dot u_k\bigl( q(t') \bigr) \,
%    \kappa_k(t - t') \, dt \, dt'
%  ,
%  \notag
%  %\label{eq:error_asym0}
%\end{equation}
%%
%with
%and
%$\kappa_k(t - t') \equiv \left\langle \theta_k(t) \, \theta_k(t') \right\rangle$.
%
%\begin{align*}
%  u_k(q')
%  &\equiv
%  e^{\lambda_k \, [q' - q(T)]},
%  \\
%  \kappa_k(t - t')
%  &\equiv
%  \left\langle
%    \theta_k(t) \, \theta_k(t')
%  \right\rangle.
%\end{align*}

@@

To model the initial error for Eq. \eqref{eq:error_res},
we will assume that
the adaptive FDS sampling
was performed under a constant updating magnitude
for sufficiently long.
%
Then, one can show that
\begin{equation}
  \left\langle y_k^2(0) \right\rangle = \frac 1 2 \, \Gamma_k \, \lambda_k \, a_0
  ,
  \label{eq:y2_eql}
\end{equation}
%
\note{
Consider a long period, $T'$, under a fixed schedule,
$\alpha(t) = a_0$,
the residual error becomes negligible, and
the $k$th component of the asymptotic error
is given by
%
\begin{align*}
  \left\langle y_k^2(T') \right\rangle
  =
  \int_0^{T'}
    \Gamma_k \, (\lambda_k \, a_0)^2 \,
      e^{ 2 \, \lambda_k \, a_0 \, (t - T') }
    \, dt
  \stackrel{ T' \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  \frac 1 2 \, \Gamma_k \, \lambda_k \, a_0
  .
\notag
%\label{eq:error_eql}
\end{align*}
}
%
and
%
\begin{equation}
  \Err_R(T)
  =
  \sum_{k = 2}^n
      \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k \,
      e^{  - 2 \, \lambda_k \, q(T) }
  .
  \label{eq:error_res1}
\end{equation}
%
This expression excludes systematic bias
and thus serves as a lower bound for a realistic simulation.

@@

Many commonly-used updating schemes,
such as the Gaussian updating scheme
used in metadynamics,
can be characterized by a rigid window
centered on the current bin.
%
For these translationally invariant or homogeneous
schemes, the eigenvalues can be readily found
from the shape of the window function.
%

%It is natural to first introduce homogeneous updating schemes
For a periodic variable\cite{dama2014},
the updating matrix, $\mathbf w$,
assumes the following form
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\notag
%\label{eq:w_band_pbc}
\end{equation}
%
where the numbers,
$\mu_{-b}, \dots, \mu_b$, ($b \le n/2$)
characterizing the rigid window
will be referred to as an \emph{updating kernel}\cite{bussi2006}.
%
To satisfy Eq. \eqref{eq:w_sumj},
we impose the normalization
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
with $\mu_l = 0$ for $|l| > b$.
%
If $b = n/2$, $\mu_{-b}$ is removed
from the sum in Eq. \eqref{eq:mu_normalization}
to avoid double counting.
%
For simplicity, we also impose the symmetry
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}
%
As shown in Appendix \ref{sec:wband_eig},
%the multiplication of $\mathbf w$
%is equivalent to a convolution with the kernel, and
the eigenvalues of $\mathbf w$ are
the Fourier or cosine transform of the updating kernel:
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2 \,
  \sum_{ l = 1 }^b
  \mu_l
  \cos
  \frac{ (k - 1) \, l \, 2 \, \pi }
       {            g \, n        }
  ,
  \label{eq:wband_eigenvalue}
\end{equation}
%
where $g = 1$ here for the periodic variable.

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
that ignore short-wavelength eigenmodes,
which are typically local noise.
%
Mathematically, these updating schemes
allow us to assign zero eigenvalues
to modes that are likely absent from the PMF.
%
For example,
%if we can assume
by assuming
that the PMF is sufficiently smooth
such that the eigenmodes can be limited to $k \le K + 1$,
%
we can construct an updating scheme with
$$
\lambda_1 = \cdots = \lambda_{K+1} = 1,
\mathrm{\; and \;}
\lambda_{K+2} = \cdots = \lambda_n = 0.
$$
%
Then, the updating matrix is given by
%
$$w_{ij} = \sum_{k=1}^{K+1} \phi_{ki} \, \phi_{kj},$$
%
and Eq. \eqref{eq:error_asym_singlebin} is modified as
%
\begin{equation}
  \Err_A(T)
  =
  \frac {       T     }
        { (T + t_0)^2 }
  \sum_{ k = 2 }^{K+1}
    \Gamma_k.
%\notag
\label{eq:error_asym_sinc}
\end{equation}
%
The optimal schedule of these schemes
remains to be the inverse-time schedule
given by Eq. \eqref{eq:alpha_invt1},
%
with
%
\begin{equation}
  t_0
  =
  \frac{        1      }
       { \alpha(t = 0) }
  =
  \frac{  2  }
       { a_0 }
  ,
\label{eq:t0_sinc}
\end{equation}
%
where we have used Eq. \eqref{eq:half_alpha0}
in the second step.
%
\note{
For the single-bin updating scheme,
it recovers Eq. \eqref{eq:t0_singlebin}
under the assumption of Eq. \eqref{eq:y2_eql}.
%
From Eq. \eqref{eq:y2_eql}, we have
$$E(0) = \frac{1}{2} a_0 \sum_{k=2}^n \Gamma_k = \frac{1}{2} \, a_0 \, \Gamma.$$
}
%
From Eqs. \eqref{eq:error_tot},
\eqref{eq:error_res1},
\eqref{eq:tk0},
and
\eqref{eq:error_asym_sinc},
we get
%
\begin{equation}
  \Err(T)
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 2 }^{ K + 1 }
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
This result recovers Eq. \eqref{eq:Emin_singlebin}
for the single-bin updating scheme:
if we set $K + 1 = n$,
%
\begin{equation}
  \sum_{k=2}^n \Gamma_k
  = \sum_{i=1}^n \rho_i \, G_i = \Gamma
  ,
  \notag
  %\label{eq:Gamma_G_sum}
\end{equation}
which can be shown by following the steps
similar to those leading to Eq. \eqref{eq:y2_sum}.
%
%
%
\note{The derivation.
From Eq. \eqref{eq:eig_orthonormal_cols}, we have
$\sum_{i=1}^n \rho_i \langle x_i(t) \, x_i(0) \rangle
=\sum_{k=2}^n \rho_i \langle y_k(t) \, y_k(0) \rangle$,
so
\begin{align*}
\sum_{i=1}^n \rho_i G_i
&=\sum_{t=-\infty}^\infty\sum_{i=1}^n \rho_i \langle x_i(t) \, x_i(0) \rangle \\
&=\sum_{t=-\infty}^\infty\sum_{k=2}^n \rho_i \langle y_k(t) \, y_k(0) \rangle
=\sum_{k=2}^n \Gamma_k.
\end{align*}
}
%
This error is roughly proportional to the inverse simulation length,
as one would expect from an equilibrium FDS simulation.
%In Appendix \ref{sec:equilerr},
%we show that the above error tends
%to the error of the ideal equilibrium FDS simulation
%in the asymptotic limit.

@@

To find the eigenvectors for a periodic variable, $\phi_i$,
we define the out-of-boundary values by
%
\begin{equation}
  \phi_j = \phi_{j \pm n},
\label{eq:phi_pbc}
\end{equation}
%
such that $j \pm n$ lies in between $1$ and $n$.
%
Then, the multiplication of the matrix, $\mathbf w$,
is equivalent to a convolution with the kernel:
%
\begin{equation}
  \sum_{ j = 1 }^n
    w_{ij} \, \phi_j
  =
  \sum_{ j = 1 - n }^{ 2 \, n }
    \mu_{i - j} \, \phi_j
  =
  \sum_{ l = -b }^{ b }
    \mu_l \, \phi_{ i - l}
  .
\label{eq:wmul_to_convol}
\end{equation}
%
\note{The derivation of the first step of
  Eq. \eqref{eq:wmul_to_convol}:
$$
\begin{aligned}
  \sum_{j = 1}^n
    w_{ij} \, \phi_j
  &=
  \sum_{j = 1}^n
    \mu_{i - j} \, \phi_j
  +
  \sum_{j = 1}^n
    \mu_{i - j - n} \, \phi_j
  +
  \sum_{j = 1}^n
    \mu_{i - j + n} \, \phi_j
  \\
  &=
  \sum_{j = 1}^n
    \mu_{i - j} \, \phi_j
  +
  \sum_{l = 1+n}^{2 \, n}
    \mu_{i - l} \, \phi_{l - n}
  +
  \sum_{l = 1-n}^0
    \mu_{i - l} \, \phi_{l + n}
  \\
  &=
  \sum_{j = 1}^n
    \mu_{i - j} \, \phi_j
  +
  \sum_{l = 1+n}^{2 \, n}
    \mu_{i - l} \, \phi_{l}
  +
  \sum_{l = 1-n}^0
    \mu_{i - l} \, \phi_{l}
  \\
  &=
  \sum_{j = 1-n}^{2 \, n}
    \mu_{i - j} \, \phi_j
  .
\end{aligned}
$$
The second step of Eq. \eqref{eq:wmul_to_convol}
follows from the constraint, $\mu_l = 0$, for $|l| > b$.

>>>>

%To simplify the ensuing discussion,
Second, we limit ourselves to matrices % $\mathbf w$'s
that satisfy
the detailed balance condition,
%
\begin{equation}
  \rho_i \, w_{ij} = \rho_j \, w_{ji}
  ,
  \label{eq:w_detailedbalance}
\end{equation}
%
for the error weighting distribution, $\pmb \rho$,
introduced in Eq. \eqref{eq:vbar_def}.
%
This requires the updating matrix, $\mathbf w$
to be symmetric
for a uniform $\pmb\rho$,
or more generally,
the scaled matrix, $\sqrt{ \rho_i/\rho_j } \, w_{ij}$,
to be symmetric.
%
This symmetry allows the diagonalization
of $\mathbf w$ with a set of
eigenvectors, $\phi_{ki}$,
%
\begin{equation}
  \sum_{i = 1}^n \phi_{ki} \, w_{ij}
  =
  \lambda_k \, \phi_{kj}
  ,
\label{eq:eig_w}
\end{equation}
%
satisfying the orthonormal conditions\cite{vankampen}:
%
\begin{align}
  \sum_{k = 1}^n
    \phi_{ki} \, \phi_{kj}^*
  &=
  \delta_{ij} \, \rho_i,
  \label{eq:eig_orthonormal_cols}
  \\
  \sum_{i = 1}^n
    \frac{ \phi_{ki} \, \phi_{li}^* }
         { \rho_i }
  &=
  \delta_{kl}
  ,
  \label{eq:eig_orthonormal_rows}
\end{align}
%
where the superscript $^*$
denotes the complex conjugate.

The above symmetry implies the existence of
a left eigenvector, $\pmb \rho$,
corresponding to the uniform right eigenvector
associated with Eq. \eqref{eq:w_sumj},
with eigenvalue $1$:
%
\begin{equation}
  \sum_{i = 1}^n \rho_i \, w_{ij}
  =
  \sum_{i = 1}^n \rho_j \, w_{ji}
  =
  \rho_j
  .
  \label{eq:w_balance}
\end{equation}
%
We will label this eigenvector as the first one,
and
%
\begin{equation}
  \phi_{1i} = \rho_i,
\label{eq:eigenmode1}
\end{equation}
%
with $\lambda_1 = 1$.
%
Clearly, Eq. \eqref{eq:eigenmode1}
satisfies the normalization condition
given by Eq. \eqref{eq:eig_orthonormal_rows}
for $k = l = 1$,
while the orthogonality for $k \ge 2$ demands
%
\begin{equation}
  \sum_{ i = 1 }^n \phi_{ki}
  =
  \sum_{ i = 1 }^n
    \frac{ \phi_{ki} \, \phi_{1i}^* }
         { \rho_i }
  =
  0
  .
\label{eq:ortho1}
\end{equation}
%
We may write the updating matrix in terms of $\phi_{ki}$'s as
%
\begin{equation}
  w_{ij}
  =
  \frac{1}{\rho_i} \sum_{k=1}^n
  \lambda_k \, \phi_{ki}^* \, \phi_{kj}
  .
  \notag
\end{equation}

@@

The error defined in Eq. \eqref{eq:error_def}
can be rewritten in terms of the eigenmodes of $\mathbf w$,
%
\begin{align}
  \Err
  \equiv
  \sum_{i = 1}^n \rho_i \,
    \left\langle x_i^2 \right\rangle
  =
  \sum_{i, j, k=1}^n \phi_{ki} \, \phi_{kj}^* \,
    \left\langle x_i \, x_j \right\rangle
  =
  \sum_{k = 1}^n
    \left\langle
      \left| y_k \right|^2
    \right\rangle
  ,
\notag
\end{align}
%
where we have used Eq. \eqref{eq:eig_orthonormal_cols}.
%
From Eq. \eqref{eq:eigenmode1},
the first eigenmode,
$y_1 = \sum_{j=1}^n \rho_j \, x_j$,
vanishes identically by Eq. \eqref{eq:rhox_sum}.
%
Thus, we can start the last sum %in Eq. \eqref{eq:y2_sum}
from $k = 2$,
%
\begin{equation}
  \Err(T)
  =
  \sum_{k = 2}^n
    \left\langle
      \left| y_k(T) \right|^2
    \right\rangle
  .
\label{eq:y2_sum}
\end{equation}

In analogy to Eq. \eqref{eq:ET_average},
%using Eqs. \eqref{eq:yt_diffeq} and \eqref{eq:y2_sum},
we can write the error after a period, $T$,
as a sum of two contributions,
%
\begin{equation}
  \Err(T)
  =
  \Err_R(T) + \Err_A(T),
  \label{eq:error_tot}
\end{equation}
%
where the residual error
(from the decay of the initial error)
is given by
%
\begin{equation}
  \Err_R(T)
  =
  \sum_{k = 2}^n
    \left\langle y_k^2(0) \right\rangle \,
    e^{ - 2 \, \lambda_k  \, q(T) },
  \label{eq:error_res}
\end{equation}
%
and the asymptotic error
is given by
%
\begin{align}
  \Err_A(T)
  &=
  \int_0^T
  \sum_{k = 2}^n
  \Gamma_k \, \dot u_k^2\bigl( q(t) \bigr) \, dt
  ,
\label{eq:error_asym}
\end{align}
%
where
%
\begin{equation}
  u_k(q') \equiv e^{\lambda_k \, [q' - q(T)]}
  ,
  \label{eq:uk_def}
\end{equation}
%
and $\Gamma_k$ is computed from
the autocorrelation function
of the histogram fluctuation,
$\kappa_k(t) \equiv \langle \theta_k^*(0) \, \theta_k(t) \rangle$,
as:
%
\begin{equation}
  \Gamma_k
  = \sum_{t = -\infty}^\infty \kappa_k(t)
  = \sum_{t = -\infty}^\infty
  \left\langle \theta_k^*(0) \, \theta_k(t) \right\rangle
  .
  \label{eq:Gamma_sum}
\end{equation}
%
%\begin{equation}
%$\Gamma_k = \int_{-\infty}^\infty \kappa_k(t) \, dt$,
%  \notag
%\label{eq:Gamma_integral}
%\end{equation}
%
Equation \eqref{eq:Gamma_sum}
essentially maps the histogram fluctuation, $\theta_k(t)$,
to the equivalent white noise
so that the theory can be applicable to
a general sampling process.
%
In this way,
the underlying sampling process
affects the error only through the
few numbers, $\Gamma_k$'s.
%
For example, we expect large values of $\Gamma_k$
from eigenmodes corresponding to
persistent long-range motion
in sampling complex or glassy systems.
%
In Appendix \ref{sec:Gamma},
we evaluate $\Gamma_k$'s in two special cases,
and show how to measure them
from an adaptive FDS simulation.
%
%\begin{equation}
%  \Err_A(T)
%  =
%  \int_0^T \int_0^T
%  \sum_{k = 2}^n
%    \dot u_k\bigl( q(t) \bigr) \,
%    \dot u_k\bigl( q(t') \bigr) \,
%    \kappa_k(t - t') \, dt \, dt'
%  ,
%  \notag
%  %\label{eq:error_asym0}
%\end{equation}
%%
%with
%and
%$\kappa_k(t - t') \equiv \left\langle \theta_k(t) \, \theta_k(t') \right\rangle$.
%
%\begin{align*}
%  u_k(q')
%  &\equiv
%  e^{\lambda_k \, [q' - q(T)]},
%  \\
%  \kappa_k(t - t')
%  &\equiv
%  \left\langle
%    \theta_k(t) \, \theta_k(t')
%  \right\rangle.
%\end{align*}

@@

To model the initial error for Eq. \eqref{eq:error_res},
we will assume that
the adaptive FDS sampling
was performed under a constant updating magnitude
for sufficiently long.
%
Then, one can show that
\begin{equation}
  \left\langle
    \left| y_k(0) \right|^2
  \right\rangle
  = \frac 1 2 \, \Gamma_k \, \lambda_k \, a_0
  ,
  \label{eq:y2_eql}
\end{equation}
%
\note{
Consider a long period, $T'$, under a fixed schedule,
$\alpha(t) = a_0$,
the residual error becomes negligible, and
the $k$th component of the asymptotic error
is given by
%
\begin{align*}
  \left\langle y_k^2(T') \right\rangle
  =
  \int_0^{T'}
    \Gamma_k \, (\lambda_k \, a_0)^2 \,
      e^{ 2 \, \lambda_k \, a_0 \, (t - T') }
    \, dt
  \stackrel{ T' \to \infty }
  { =\joinrel=\joinrel=\joinrel= }
  \frac 1 2 \, \Gamma_k \, \lambda_k \, a_0
  .
\notag
%\label{eq:error_eql}
\end{align*}
}
%
and
%
\begin{equation}
  \Err_R(T)
  =
  \sum_{k = 2}^n
      \frac 1 2 \, a_0 \, \Gamma_k \, \lambda_k \,
      e^{  - 2 \, \lambda_k \, q(T) }
  .
  \label{eq:error_res1}
\end{equation}
%
This expression excludes systematic bias
and thus serves as a lower bound for a realistic simulation.

@@

Many commonly-used updating schemes,
such as the Gaussian updating scheme
used in metadynamics,
can be characterized by a rigid window
centered on the current bin.
%
For these translationally invariant or homogeneous
schemes, the eigenvalues can be readily found
from the window function ($\mu_0, \mu_1, \dots, \mu_b$)
as a cosine transform,
%
\begin{equation}
  \lambda_k
  =
  \mu_0
  +
  2 \,
  \sum_{ l = 1 }^b
  \mu_l
  \cos
  \frac{ (k - 1) \, l \, 2 \, \pi }
       {            g \, n        }
  ,
  \label{eq:wband_eigenvalue}
\end{equation}
%
where $g = 1$ here for the periodic variable,
or $2$ for a non-periodic one
[cf. Appendix \ref{sec:more_wband}].

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
that ignore short-wavelength eigenmodes,
which are typically local noise.
%
Mathematically, these updating schemes
allow us to assign zero eigenvalues
to modes that are likely absent from the PMF.
%
For example,
%if we can assume
by assuming
that the PMF is sufficiently smooth
such that the eigenmodes can be limited to $k \le K + 1$,
%
we can construct an updating scheme with
$$
\lambda_1 = \cdots = \lambda_{K+1} = 1,
\mathrm{\; and \;}
\lambda_{K+2} = \cdots = \lambda_n = 0.
$$
%
Then, the updating matrix is given by
%
$$w_{ij} = n \sum_{k=1}^{K+1} \phi_{ki}^* \, \phi_{kj},$$
%
and Eq. \eqref{eq:error_asym_singlebin} is modified as
%
\begin{equation}
  \Err_A(T)
  =
  \frac {       T     }
        { (T + t_0)^2 }
  \sum_{ k = 2 }^{K+1}
    \Gamma_k.
%\notag
\label{eq:error_asym_sinc}
\end{equation}
%
The optimal schedule of these schemes
remains to be the inverse-time schedule
given by Eq. \eqref{eq:alpha_invt1},
%
with
%
\begin{equation}
  t_0
  =
  \frac{        1      }
       { \alpha(t = 0) }
  =
  \frac{  2  }
       { a_0 }
  ,
\label{eq:t0_sinc}
\end{equation}
%
where we have used Eq. \eqref{eq:half_alpha0}
in the second step.
%
\note{
For the single-bin updating scheme,
it recovers Eq. \eqref{eq:t0_singlebin}
under the assumption of Eq. \eqref{eq:y2_eql}.
%
From Eq. \eqref{eq:y2_eql}, we have
$$E(0) = \frac{1}{2} a_0 \sum_{k=2}^n \Gamma_k = \frac{1}{2} \, a_0 \, \Gamma.$$
}
%
From Eqs. \eqref{eq:error_tot},
\eqref{eq:error_res1},
\eqref{eq:tk0},
and
\eqref{eq:error_asym_sinc},
we get
%
\begin{equation}
  \Err(T)
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 2 }^{ K + 1 }
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
This result recovers Eq. \eqref{eq:Emin_singlebin}
for the single-bin updating scheme:
if we set $K + 1 = n$,
%
\begin{equation}
  \sum_{k=2}^n \Gamma_k
  = \sum_{i=1}^n \rho_i \, G_i = \Gamma
  ,
  \notag
  %\label{eq:Gamma_G_sum}
\end{equation}
which can be shown by following the steps
similar to those leading to Eq. \eqref{eq:y2_sum}.
%
%
%
\note{The derivation.
From Eq. \eqref{eq:eig_orthonormal_cols}, we have
$\sum_{i=1}^n \rho_i \langle x_i(t) \, x_i(0) \rangle
=\sum_{k=2}^n \rho_i \langle y_k(t) \, y_k(0) \rangle$,
so
\begin{align*}
\sum_{i=1}^n \rho_i G_i
&=\sum_{t=-\infty}^\infty\sum_{i=1}^n \rho_i \langle x_i(t) \, x_i(0) \rangle \\
&=\sum_{t=-\infty}^\infty\sum_{k=2}^n \rho_i \langle y_k(t) \, y_k(0) \rangle
=\sum_{k=2}^n \Gamma_k.
\end{align*}
}
%
This error is roughly proportional to the inverse simulation length,
as one would expect from an equilibrium FDS simulation.
%In Appendix \ref{sec:equilerr},
%we show that the above error tends
%to the error of the ideal equilibrium FDS simulation
%in the asymptotic limit.

@@

It is natural to first introduce homogeneous updating schemes
for a periodic variable\cite{dama2014},
the updating matrix, $\mathbf w$,
assumes the following form
%
\begin{equation}
  w_{ij}
  =
  \mu_{i-j}
  +
  \mu_{i-j+n}
  +
  \mu_{i-j-n}
  ,
\notag
%\label{eq:w_band_pbc}
\end{equation}
%
where the numbers,
$\mu_{-b}, \dots, \mu_b$, ($b \le n/2$)
characterizing the rigid window
will be referred to as an \emph{updating kernel}\cite{bussi2006}.
%
To satisfy Eq. \eqref{eq:w_sumj},
we impose the normalization
%
\begin{equation}
  \mu_{-b} + \cdots + \mu_b = 1
  ,
\label{eq:mu_normalization}
\end{equation}
%
with $\mu_l = 0$ for $|l| > b$.
%
If $b = n/2$, $\mu_{-b}$ is removed
from the sum in Eq. \eqref{eq:mu_normalization}
to avoid double counting.
%
For simplicity, we also impose the symmetry
%
\begin{equation}
  \mu_i = \mu_{-i}
  .
\label{eq:mu_symm}
\end{equation}

To find the eigenvectors,
we define for a periodic variable, $\phi_i$,
the out-of-boundary values by
%
\begin{equation}
  \phi_j = \phi_{j \pm n},
\label{eq:phi_pbc}
\end{equation}
%
such that $j \pm n$ lies in between $1$ and $n$.
%
Then, the multiplication of the matrix, $\mathbf w$,
is equivalent to a convolution with the kernel:
%
\begin{equation}
  \sum_{ j = 1 }^n
    w_{ij} \, \phi_j
  =
  \sum_{ j = 1 - n }^{ 2 \, n }
    \mu_{i - j} \, \phi_j
  =
  \sum_{ l = -b }^{ b }
    \mu_l \, \phi_{ i - l}
  .
\label{eq:wmul_to_convol}
\end{equation}
%
\note{The derivation of the first step of
  Eq. \eqref{eq:wmul_to_convol}:
$$
\begin{aligned}
  \sum_{j = 1}^n
    w_{ij} \, \phi_j
  &=
  \sum_{j = 1}^n
    \mu_{i - j} \, \phi_j
  +
  \sum_{j = 1}^n
    \mu_{i - j - n} \, \phi_j
  +
  \sum_{j = 1}^n
    \mu_{i - j + n} \, \phi_j
  \\
  &=
  \sum_{j = 1}^n
    \mu_{i - j} \, \phi_j
  +
  \sum_{l = 1+n}^{2 \, n}
    \mu_{i - l} \, \phi_{l - n}
  +
  \sum_{l = 1-n}^0
    \mu_{i - l} \, \phi_{l + n}
  \\
  &=
  \sum_{j = 1}^n
    \mu_{i - j} \, \phi_j
  +
  \sum_{l = 1+n}^{2 \, n}
    \mu_{i - l} \, \phi_{l}
  +
  \sum_{l = 1-n}^0
    \mu_{i - l} \, \phi_{l}
  \\
  &=
  \sum_{j = 1-n}^{2 \, n}
    \mu_{i - j} \, \phi_j
  .
\end{aligned}
$$
The second step of Eq. \eqref{eq:wmul_to_convol}
follows from the constraint, $\mu_l = 0$, for $|l| > b$.

Similarly,
for the left eigenvector, we have
$$
  \sum_{ i = 1 }^n
    \phi_i \, w_{ij}
  =
  \sum_{ l = -b }^b
    \phi_{j - l} \, \mu_{-l}
  .
$$
But due to the symmetry Eq. \eqref{eq:mu_symm},
it is identical to \eqref{eq:wmul_to_convol}.
}
%
As a result, the characteristic equation,
$\mathbf w \, \pmb\phi = \lambda \, \pmb\phi$,
can be solved by discrete Fourier transform.
%
We may construct a set of orthonormal eigenvectors,
$\pmb\phi^{(1)}, \dots, \pmb\phi^{(n)}$,
compatible with the periodicity, Eq. \eqref{eq:phi_pbc},
%
\begin{equation}
  \phi^{(k)}_j
  =
  \phi_{kj}
  =
  \frac 1 n
  \exp
  \left[
    \ii
    \frac{ (k - 1) \, j \, 2 \, \pi }
         {            n             }
  \right]
  ,
  \notag
\end{equation}
%
where $\ii^2 = -1$.
%
Alternatively, to avoid complex numbers, we can use
%
\begin{equation}
  \phi^{(k)}_j
  =
  \phi_{kj}
  =
  \frac{ \sqrt{ 2 - \delta_{k,0} } } { n }
  %\overset{ \cos } { \sin }
  \,
  {\cos \atop \sin}
  \left[
    \frac{ (k - 1) \, j \, 2 \, \pi }
         {            n             }
  \right]
  ,
  \notag
\end{equation}
%
where the function takes the cosine form for $k \le n/2 + 1$,
or the sine form otherwise.
%
Then one can readily verify that the eigenvalues are given by
  Eq. \eqref{eq:wband_eigenvalue},
  and there is a two-fold degeneracy,
  $\lambda_{n - k + 2} = \lambda_k$.
\note{Consider the unnormalized version
  $$
  \Phi^{(k)}_j =
  \exp\left[
    \frac{ ( k - 1 ) \, j \, 2 \, \pi }
         {              n             }
    \ii
  \right]
  ,
  $$
  we have
  $$
  \begin{aligned}
  \sum_{j = 1}^n
    w_{ij} \, \Phi^{(k)}_j
  &=
  \sum_{l = -b}^b
    \mu_l \, \Phi^{(k)}_{i - l}
  \\
  &=
  \mu_0 \, \Phi^{(k)}_i
  +
  \sum_{l = 1}^b
    \mu_l \,
    \left[ \Phi^{(k)}_{i - l} + \Phi^{(k)}_{i + l} \right]
  \\
  &=
  \Phi^{(k)}_i \,
  \left[
    \mu_0
    +
    2 \sum_{l = 1}^b
      \mu_l \, \cos
      \frac{ (k - 1) \, l \, 2 \, \pi }
           {            n             }
  \right]
  .
  \end{aligned}
  $$
  The term in the square brackets is the eigenvalue given by
  Eq. \eqref{eq:wband_eigenvalue}.
}
%
